A Tale of Two Transects

### 2020-02-12 writing/transect_paper.Rmd
work on results, knit versions to send before we work together on the discussion. Also send more figure code to the plan.

---
title: "A Tale of Two Transects"
subtitle: "Also Two Sequencing Technologies, and Two Amplicons"
author: 
  - Brendan Furneaux^1^,
    Mohammad Bahram^1^,
    Anna Rosling^3^,
    Nourou S. Yorou^2^,
    Martin Ryberg^1^
  - ^1^Program in Systematic Biology,
    Department of Organismal Biology,
    Uppsala University,
    Uppsala, Sweden
  - ^2^Research Unit in Tropical Mycology and Plant-Fungi Interactions,
    LEB,
    University of Parakou,
    Parakou, Benin
  - ^3^Program in Evolutionary Biology,
    Department of Ecology and Genetics,
    Uppsala University,
    Uppsala, Sweden
date: "`r format(Sys.time(), '%B %d, %Y')`"
bibliography:
  - "all.bib"
  - "R.bib"
csl: molecular-ecology.csl
always_allow_html: true
output:
  bookdown::pdf_document2:
    fig_crop: yes
    keep_tex: yes
    latex_engine: lualatex
    includes:
      in_header: preamble.tex
  bookdown::word_document2: 
    keep_md: yes
---

@@

Sampling was conducted at two sites (Ang: N 9.75456° W 2.14064°; Gan: N 9.75678° W 2.31058 °) approximately 30 km apart in the _Forêt Reservée de l'Ouémé Supérieur_ (Upper Ouémé Forest Reserve) in central Benin.
Both locations were located in woodlands dominated by the ECM host tree _Isoberlinia doka_ (Caesalpinioideae).
At each site, 25 soil samples were collected along a linear transect at intervals of 1 m in May 2015.
One third of the sample locations (3-m spacing) were resampled one year later in June 2016.
For each sample, any coarse organic debris <!--Do you have any definition? Wider than 5 mm? MR--> was removed from the soil surface and a sample of approximately 5cm×5cm×5cm was extracted with a <!-- sterilized? -MR--> knife blade.
Each sample was sealed in a plastic zipper bag and homogenized by shaking and manually breaking apart soil aggregations.
Approximately 50 mg total of soil was collected from two locations in the homogenized soil sample and placed into a separate 2.0 mL microtube containing 750 mL of lysis buffer and lysis beads (Xpedition^TM^ Soil/Fecal DNA miniprep, Zymo Research Corporation, Irvine, California, USA) and lysed in the field using a handheld bead-beater (TerraLyser^TM^; Zymo Research Corporation).
<!--The remainder of each sample was dried at 40°C with an electric dehydrator within 24-48 hours of collection for bulk chemical analysis. -->

@@

Forward and reverse Illumina reads were quality controlled (truncation after first base with quality <= 10, and then max 3 expected errors in either read) and denoised using `DADA2`.
Forward and reverse reads were denoised using seperate error models and then merged.

We attempted to denoise both long and short PacBio amplicons using `DADA2` version `r packageVersion("dada2")` [@callahan2016; @callahan2019] according to the steps outlined in  the supplementary information in @callahan2019.
However, only `r readd(region_table)$ASVs[readd(region_table)$region=="long"]` amplicon sequence variants (ASVs) were obtained for the long amplicons, representing `r percent(readd(region_table)$reads[readd(region_table)$region == "long"]/as.integer(gsub(" ", "", readd(demuxcounts)["pb_500"])), format = "d")` of the trimmed reads.
We believe that this poor performance was due to a combination of long read length and low sequencing depth relative to community diversity.
The `DADA2` algorithm requires that the seed sequence of each ASV be represented by at least two error-free reads [@callahan2016].
If sequencing errors are uniformly distributed, then the probability that a given read will be error-free is $(1-\epsilon)^L$, where $\epsilon$ is the sequencing error rate and $L$ is the read length in base pairs.
Then the number of reads of a given sequence that would be required to obtain two error-free reads in expectation is $2/(1-\epsilon)^L$.
For the combination of long reads (median $L=`r readd(demuxlength)['pb_500']` \text{ bp}$ after trimming) and moderate error rate (mean $\epsilon= `r format(readd(demuxqual)['pb_500'], digits = 4)`$ based on ccs quality scores) for the long amplicon in this study, the expected number of reads required to achieve two error-free reads is `r format(as.integer(round(2/(1 - readd(demuxqual)['pb_500'])^readd(demuxlength)['pb_500'])), big.mark = " ")`.
Given the high diversity relative to sequencing depth in this study (`r readd(region_table)$ASVs[readd(region_table)$seq_run == "pb_483" & readd(region_table)$region == "short"]` ASVs based on PacBio short amplicons, `r readd(demuxcounts)["pb_500"]` trimmed long amplicon reads), this requirement could not have been met for the long amplicons except by the most abundant sequences.
In comparison, the equivalent requirement for the short amplicon ($L=`r readd(demuxlength)['pb_483']` \text{ bp}$, $\epsilon=`r readd(demuxqual)['pb_483']`$) is only `r round(2/(1 - readd(demuxqual)['pb_483'])^readd(demuxlength)['pb_483'], 1)` reads.
We therefore developed a new workflow to assemble ASVs from the long amplicons, as follows:

@@

The number of sequencing reads and ASVs at each stage in the bioinformatics pipeline are shown in Table \@ref(tab:bioinfo).
Long (short) amplicon sequencing with PacBio yielded `r readd(rawcounts)["pb_500"]` (`r readd(rawcounts)["pb_483"]`) circular consensus reads, of which `r readd(demuxcounts)["pb_500"]` (`r readd(demuxcounts)["pb_483"]`) were successfully demultiplexed.
The 5.8S region was detected in `r readd(regioncounts)["pb_500"]` (`r readd(regioncounts)["pb_483"]`) reads.
Ion Torrent sequencing yielded `r readd(demuxcounts)["is_057"]` demultiplexed reads, with the 5.8S region detected in `r readd(regioncounts)["is_057"]`.

@@

Figure \@ref(fig:pb-vs-it) shows the correspondence between the read count for different ASVs/OTUs in the different technologies, where shared ASVs/OTUs are plotted as circles, and unshared OTUs are plotted on the margins.
In all cases, the read counts for shared ASVs/OTUs were correlated ($R^2 = `r compR2("SH-2257", "is_057", "OTU")`$).
Correlations between read counts for the three technologies using the short amplicon library were increased by OTU clustering
(`r compR2("pb_483", "SH-2257", "ASV")` to `r compR2("pb_483", "SH-2257", "OTU")`,
`r compR2("pb_483", "is_057", "ASV")` to `r compR2("pb_483", "is_057", "OTU")`, and
`r compR2("SH-2257", "is_057", "ASV")` to `r compR2("SH-2257", "is_057", "OTU")`,
for PacBio vs. Illumina, PacBio vs. Ion Torrent, and Illumina vs. Ion Torrent, respectively),
but this was not, or only weakly, the case for correlations between the long amplicon library and short amplicon library
(`r compR2("pb_500", "pb_483", "ASV")` to `r compR2("pb_500", "pb_483", "OTU")`,
`r compR2("pb_500", "SH-2257", "ASV")` to `r compR2("pb_500", "SH-2257", "OTU")`, and
`r compR2("pb_500", "is_057", "ASV")` to `r compR2("pb_500", "is_057", "OTU")`, for PacBio long amplicon reads vs PacBio, Illumina, and Ion Torrent short reads, respectively).

@@

## Spatio-temporal variogram

Variograms relating community dissimilarities and spatial distances are shown in Figure \@ref(fig:spatio).
For the cases where the Mantel correlogram detected a significant positive correlation, the best-fit exponential curve is also displayed; for the long amplicon library, where there was also a significant positive correlation after a time lag of one year, the best fit curve for time lag is also included.
The best fit spatial and temporal ranges, along with a 95% confidence interval, are shown in \@ref(tab:variofit).
Although a model was fit for the weighted UNIFRAC distance applied to the total fungal community, the result was very poorly constrained, and a range of 0m was included in the 95% confidence interval.
Although the best fit spatial ranges vary between the different datasets by a factor of about 3, there is overlap of the 95% confidence intervals for all of the Bray-Curtis variograms across communities, amplicon libraries, and sequencing technologies.

>>>>

---
title: "A Tale of Two Transects"
subtitle: "Also Two Sequencing Technologies, and Two Amplicons"
author: 
  - Brendan Furneaux^1^,
    Mohammad Bahram^1^,
    Anna Rosling^3^,
    Nourou S. Yorou^2^,
    Martin Ryberg^1^
  - ^1^Program in Systematic Biology,
    Department of Organismal Biology,
    Uppsala University,
    Uppsala, Sweden
  - ^2^Research Unit in Tropical Mycology and Plant-Fungi Interactions,
    LEB,
    University of Parakou,
    Parakou, Benin
  - ^3^Program in Evolutionary Biology,
    Department of Ecology and Genetics,
    Uppsala University,
    Uppsala, Sweden
date: "`r format(Sys.time(), '%B %d, %Y')`"
bibliography:
  - "all.bib"
  - "R.bib"
csl: molecular-ecology.csl
always_allow_html: true
output:
  bookdown::word_document2: 
    keep_md: yes
  bookdown::pdf_document2:
    fig_crop: yes
    keep_tex: yes
    latex_engine: lualatex
    includes:
      in_header: preamble.tex
---

@@

Sampling was conducted at two sites (Ang: N 9.75456° W 2.14064°; Gan: N 9.75678° W 2.31058 °) approximately 30 km apart in the _Forêt Reservée de l'Ouémé Supérieur_ (Upper Ouémé Forest Reserve) in central Benin.
Both locations were located in woodlands dominated by the ECM host tree _Isoberlinia doka_ (Caesalpinioideae).
At each site, 25 soil samples were collected along a linear transect at intervals of 1 m in May 2015.
One third of the sample locations (3 m spacing) were resampled one year later in June 2016.
For each sample, any coarse organic debris <!--Do you have any definition? Wider than 5 mm? MR--> was removed from the soil surface and a sample of approximately 5cm×5cm×5cm was extracted with a <!-- sterilized? -MR--> knife blade.
Each sample was sealed in a plastic zipper bag and homogenized by shaking and manually breaking apart soil aggregations.
Approximately 50 mg total of soil was collected from two locations in the homogenized soil sample and placed into a separate 2.0 mL microtube containing 750 mL of lysis buffer and lysis beads (Xpedition^TM^ Soil/Fecal DNA miniprep, Zymo Research Corporation, Irvine, California, USA) and lysed in the field using a handheld bead-beater (TerraLyser^TM^; Zymo Research Corporation).
<!--The remainder of each sample was dried at 40°C with an electric dehydrator within 24-48 hours of collection for bulk chemical analysis. -->

@@

Forward and reverse Illumina reads were truncated after the first base with quality score <= 10, and then reads with more than 3 expected errors in either read were discarded.
Forward and reverse reads were denoised using `DADA2` version `r packageVersion("dada2")` [@callahan2016] using seperate error models and then merged.

We attempted to denoise both long and short PacBio amplicons using `DADA2` according to the steps outlined in  the supplementary information in @callahan2019.
However, only `r readd(region_table)$ASVs[readd(region_table)$region=="long"]` amplicon sequence variants (ASVs) were obtained for the long amplicons, representing `r percent(readd(region_table)$reads[readd(region_table)$region == "long"]/as.integer(gsub(" ", "", readd(demuxcounts)["pb_500"])), format = "d")` of the trimmed reads.
We believe that this poor performance was due to a combination of long read length and low sequencing depth relative to community diversity.
The `DADA2` algorithm requires that the seed sequence of each ASV be represented by at least two error-free reads [@callahan2016].
If sequencing errors are uniformly distributed, then the probability that a given read will be error-free is $(1-\epsilon)^L$, where $\epsilon$ is the sequencing error rate and $L$ is the read length in base pairs.
Then the number of reads of a given sequence that would be required to obtain two error-free reads in expectation is $2/(1-\epsilon)^L$.
For the combination of long reads (median $L=`r readd(demuxlength)['pb_500']` \text{ bp}$ after trimming) and moderate error rate (mean $\epsilon= `r format(readd(demuxqual)['pb_500'], digits = 4)`$ based on ccs quality scores) for the long amplicon in this study, the expected number of reads required to achieve two error-free reads is `r format(as.integer(round(2/(1 - readd(demuxqual)['pb_500'])^readd(demuxlength)['pb_500'])), big.mark = " ")`.
Given the high diversity relative to sequencing depth in this study (`r readd(region_table)$ASVs[readd(region_table)$seq_run == "pb_483" & readd(region_table)$region == "short"]` ASVs based on PacBio short amplicons, `r readd(demuxcounts)["pb_500"]` trimmed long amplicon reads), this requirement could not have been met for the long amplicons except by the most abundant sequences.
In comparison, the equivalent requirement for the short amplicon ($L=`r readd(demuxlength)['pb_483']` \text{ bp}$, $\epsilon=`r readd(demuxqual)['pb_483']`$) is only `r round(2/(1 - readd(demuxqual)['pb_483'])^readd(demuxlength)['pb_483'], 1)` reads.
We therefore developed a new workflow to assemble ASVs from the long amplicons, as follows:

@@

The number of sequencing reads and ASVs at each stage in the bioinformatics pipeline are shown in Table \@ref(tab:bioinfo).
Long (short) amplicon sequencing with PacBio yielded `r readd(rawcounts)["pb_500"]` (`r readd(rawcounts)["pb_483"]`) circular consensus reads, of which `r readd(demuxcounts)["pb_500"]` (`r readd(demuxcounts)["pb_483"]`) were successfully trimmed and demultiplexed.
Illumina sequencing yielded `r readd(rawcounts)["SH-2257"]` read pairs, of which `r readd(demuxcounts)["SH-2257"]` were successfully trimmed and demultiplexed, and `r readd(filtercounts_full)["SH-2257"]` passed quality filtering.
Ion Torrent sequencing yielded `r readd(demuxcounts)["is_057"]` demultiplexed reads.

@@

Figure \@ref(fig:pb-vs-it) shows the correspondence between the read count for different ASVs/OTUs in the different technologies, where shared ASVs/OTUs are plotted as circles, and unshared OTUs are plotted on the margins.
In all cases, the read counts for shared ASVs/OTUs were correlated ($R^2 = `r compR2("SH-2257", "is_057", "OTU")` to $).
Correlations between read counts for the three technologies using the short amplicon library were increased by OTU clustering
(`r compR2("pb_483", "SH-2257", "ASV")` to `r compR2("pb_483", "SH-2257", "OTU")`,
`r compR2("pb_483", "is_057", "ASV")` to `r compR2("pb_483", "is_057", "OTU")`, and
`r compR2("SH-2257", "is_057", "ASV")` to `r compR2("SH-2257", "is_057", "OTU")`,
for PacBio vs. Illumina, PacBio vs. Ion Torrent, and Illumina vs. Ion Torrent, respectively),
but this was not, or only weakly, the case for correlations between the long amplicon library and short amplicon library
(`r compR2("pb_500", "pb_483", "ASV")` to `r compR2("pb_500", "pb_483", "OTU")`,
`r compR2("pb_500", "SH-2257", "ASV")` to `r compR2("pb_500", "SH-2257", "OTU")`, and
`r compR2("pb_500", "is_057", "ASV")` to `r compR2("pb_500", "is_057", "OTU")`, for PacBio long amplicon reads vs PacBio, Illumina, and Ion Torrent short reads, respectively).

@@

Distance-decay curves relating community dissimilarities and spatial distances are shown in Figure \@ref(fig:spatio).
For the cases where the Mantel correlogram detected a significant positive correlation, the best-fit exponential curve is also displayed; for the long amplicon library, where there was also a significant positive correlation after a time lag of one year, the best fit curve for time lag is also included.
The best fit spatial and temporal ranges, along with a 95% confidence interval, are shown in \@ref(tab:variofit).
Although a model was fit for the weighted UNIFRAC distance applied to the total fungal community, the result was very poorly constrained, and a range of 0 m was included in the 95% confidence interval.
Although the best fit spatial ranges vary between the different datasets by a factor of about 3, there is overlap of the 95% confidence intervals for all of the Bray-Curtis variograms across communities, amplicon libraries, and sequencing technologies.
